---
title: "Homework 1"
author: Peter Sullivan

output: 
  prettydoc::html_pretty:
    theme: cayman
---

What is the association between education and earnings?  Using the 1990 dataset entitled "Work, Family, and Well-being in the United States" (https://github.com/avehtari/ROS-Examples/tree/master/Earnings/data, please do the following in this markdown document:

1) read in data found in earnings.csv;
2) graph the data and add a fitted line;
3) fit a linear regression of earnings with education as a predictor;
4) explain what each of the following represents and how it was calculated (see Lab 1c as a reference);
  a) b1hat for education
  b) b0hat 
  c) yhat
  d) uhat; and
5) interpret the estimated coefficient for education on earnings as well as the R-Squared.

#  1. Reading in The DAtA

```{r Load Data, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rstanarm)
setwd("C:/Users/pjsul/OneDrive/Desktop/DACS603/HW1")

```


```{r}

urlfile <- "https://raw.githubusercontent.com/avehtari/ROS-Examples/master/Earnings/data/earnings.csv"

data <- read_csv(urlfile, na = c("",NA))

sapply(data, function(Count) sum(is.na(Count)))
```


# 2. Graph the Data and Add a Fitted Line

I've decided to look into the correlation between weight and earnings. Weight will be on the x axis, and earnings on the Y.


```{r}
#colnames(data)

fit <- lm(earn ~ weight, data = data)

summary(fit)


plot(data$weight,data$earn, ylim = c(0,1E5))+abline(fit)


```

# 3. fit linear regression with education as a indicator

```{r}
#colnames(data)

fit <- lm(earn ~ education, data = data)

summary(fit)


plot(data$education,data$earn, ylim = c(0,2E5))+abline(fit)



```
# 4  Explainations:
explain what each of the following represents and how it was calculated (see Lab 1c as a reference);

  a) b1hat for education
  
 B1 hat =  2709.7 
 
  For every 1 year in education, earnings go up 2709.7 dollars, This is the slop of the regression line.
  
  b) b0hat 
  
  B0 hat = -14,724.1 
  
  If you had 0 years of education, you would have earned $-14,724. This is the Intercept of the regression line.
  
  c) 
  
  yhat  = -14,724 + 2709 *X 
  Yhat is the expected value of y given x using our regression model. This is the fitted model using the lm regression.
  
  d) uhat
  
  Uhat represents the residuals from the actual data to the residuals. It can be seen as the reported error from the model per data point.
  Uhat is the variance from each point the line of best fit. The mean of uhat is basically zero, as we would expect.
  
```{r}
uhat <- resid(fit)
mean(uhat)
```
  
  
E: R^2

Multiple R-squared:  0.09445

9.45 % of the variance in the model is explained by the line of best fit. This is a poor model, and I would not recommend using it.


# Looking into other variables


```{r results='hide'}


fit1 <- stan_glm(data$earn~data$education)
summary(fit1)

fit2 <- lm(data$earn~ data$age)
summary(fit2)



fit3 <- lm(data$earn~ data$height)
summary(fit3)




fit4 <- lm(data$earn~ data$father_education)
summary(fit4)


```

```{r echo = FALSE,message=FALSE, warning=FALSE, error=FALSE}
par(mfrow = c(2,2))
plot(data$education,data$earn)+abline(fit1)
plot(data$age,data$earn)+abline(fit2)
plot(data$height, data$earn)+abline(fit3)
plot(data$father_education,data$earn)+abline(fit4)


```
  
# Looping through the models
  
  
  
```{r}
library(knitr)
data

#corr_data <- data[,c(1,2,4,5,7,9:15)]
#var(corr_data, na.rm = FALSE)
#kable(round(cor(corr_data),2))

columns <- as.list(colnames(data))



models <- lapply(paste("earn ~", columns), as.formula)
models
for (model in models){
  fit <- lm(model, data = data)
  x <-summary(fit)
  print(paste(format(model),"R^2 value: ",round(x$r.squared,3)*100,"%"))
}


```
  
